---
output: html_document
---

# Dump Chapter {#ch14}


::: {.infobox .hazard data-latex="hazard"}
PRACTICE SAFE STATISTICS

Don't drink and derive
:::





# Introduction {-}
<!-- # Introduction {.para1 -} <- this adds para1 class to introduction--> 

First note - Having `{-}` in the heading skips numbering assignment (which can be customized in _bookdown.yml), so that `Chapter 1` doesn't read  as `2 Chapter 1`

<p class="para1">this is a test</p>
<div class="para1"> hello </div>

This is a cool book that we are going to make, and it's going to have information about statistics

index.Rmd is the first file loaded, similar to HTML use of index.html

Do note that

- This book uses markdown, so we can use *markdown* **notation** when writing, or <u> html tags</u> 
  + That includes sublists too
    - forever
      * but not too many
- It can be published in a lot of ways, but [this book](https://bookdown.org/yihui/bookdown/) suggests we focus on HTML first since pdf can be wonky and change format frequently
- We can use math too if we want. sex  $\rightarrow$ $\int e^X$

Since this is done in Rstudio with markdown, we can build changes by just pushing Knit at the top (Ctrl+Shift+K !). However, if we add a new file/chapter, we have to, from the console, input

```{r, eval = FALSE}
library(magrittr)
bookdown::render_book('index.Rmd', 'bookdown::gitbook')
```
(or whichever format we want. When closer to finshed, we can make a script to compile finished versions in multiple formats)

Things you'll need:

```{r, eval=FALSE}
install.packages(c("knitr", "bookdown", "servr", "shiny", "rmarkdown"))
```


https://www.w3schools.com/css/css_selectors.asp

https://bookdown.org/yihui/rmarkdown-cookbook/custom-blocks.html

## Stuff lifted that used to be in ch2 (may or may not be useful still?)


We can also have R code

```{r sampHist, include = FALSE}
normPop <- rnorm(1e5, mean = 10, sd = 20)

x <- sample(normPop, size = 5, replace = TRUE)
y <- sample(normPop, size = 100, replace = TRUE)

hist(normPop)
abline(v = 10, col = "blue", lwd = 2)
abline(v = mean(x), col = 'red', lty = 2, lwd = 2)
abline(v = mean(y), col = 'green', lty = 2, lwd = 2)
```

But that's boring and dumb when we can also have Shiny, though a few things of note, I guess

- While rmarkdown can host a self contained shiny app, bookdown requires a url. This is not ideal, and I will investigate if there is something that can be done about this
- Ok, whatever, I'm sure the library will host it for us. That would require internet access, though.
- We can still include the actual directories in our repo, or I think we can embed shiny if it is created as a single document instead of chapters
- This is dumb, but the shiny.rstudio website stopped hosting all of their shiny examples (?), and everything else is too complicated
- Oh, I know, we can just use the shiny server for public health!
- "To protect your security, ph-shiny.iowa.uiowa.edu will not allow Firefox to display the page if another site has embedded it. To see this page, you need to open it in a new window."
- Here is a way complicated one that takes too long to load. Things I don't like:
  - Small window (maybe can adjust?)
  - Scrolly window
  - Yihui recommends miniUI for embedded shiny apps, but still
  - Not yet sure what other options might exist




## Shiny test


## Regular UI
<!-- frameBorder="0" -->
<iframe src="https://ph-shiny.iowa.uiowa.edu/collin/textbook/confidenceIntervals/" width = "800" height = "750"> 
</iframe>

## Editing text

```diff
Here are some things that were written that seem fine
- this is maybe something we're not sure about
+ this is an idea or proposal we can discuss?
```

But the actual code above looks like this


    ```diff
    Here are some things that were written that seem fine
    - this is maybe something we're not sure about
    + this is an idea or proposal we can discuss?
    ```


## Idea

Not sure where else to write this - for running a simulation at home. Consider heating a pan, throwing on 10 corn kernels, and figuring out how many have popped after certain number of time. Pretend 3 minutes is 50%. Well, have them do ten for 3 minutes, then count how many popped. Repeat.


## CLT Stuff

https://iwant2study.org/lookangejss/math/ejss_model_GaltonBoardwee/GaltonBoardwee_Simulation.xhtml

Ideas to make things CLT interesting - The Wall TV show.



Questions we can answer at the end of this chapter

1. What is a statistic? What is a parameter? What is the relationship between the two? (Draw a picture would be a cool exercise. Idk how you'd grade it)
2. Why are sampling distributions random? Suppose we wanted to ask the state of Iowa which 
university had the best football team, and we only ask in Iowa City. Would this be a 
random sample? Could we use this sample to make reliable predictions about the entire state? Why or why not?
3. Prove the Central Limit Theorem using characteristic functions for the iid case, assuming finite mean and variance.

